---
title: "R Notebook"
output: html_notebook
author: Aaron J. Wilk
---

Integrative analysis of peripheral immune dynamics in SARS-CoV-2 infection
Please see our publication at: https://www.nature.com/articles/s41591-020-0944-y

This notebook will enable generation of the Seurat object used for post-processing from count matrices deposited on GEO. For code on downstream analysis and figure reproduction, please see the notebook "covid_analysis_markdown.Rmd"

Load libraries
  # NOTE: Please be sure to run the script "additionalFunctions_repo.R" to read homemade functions into the global environment

```{r}
library(Matrix)
library(Matrix.utils)
library(plyr)
library(dplyr)
library(Seurat)
library(sctransform)
library(igraph)
library(factoextra)
library(ComplexHeatmap)
library(circlize)
library(EpicTools)
require(Hmisc)
require(dplyr)
require(openxlsx)
require(ggplot2)
library(ggpubr)
require(cowplot)
library(data.table)
library(RColorBrewer)
library(rowr)
library(SingleR)
library(scater)
library(pheatmap)
library(nichenetr)
library(tidyverse)
```

# LOADING DATA #
We sequenced non-cryopreserved PBMCs from 14 donors. 8 from the ICU or floor with confirmed COVID-19 and 6 healthy age-matched controls. 

Download built count matrices from GEO (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150728) and save them in the same directory. 

Additionally, download the count matrix for patient 560 (AKA C6) from the github repository (https://github.com/ajwilk/2020_Wilk_COVID/blob/master/data/count_matrices/560_cell.counts.matrices.rds?raw=true). This patient did not consent to the sharing of sequence-level data and thus we are only able to share processed data and there is no GEO entry. 

```{r load data}
path = "path_to_count_matrices_from_GEO"
cm.list = paste0(path, list.files(pattern = "*.matrices.rds", path = path))
cm.files <- lapply(cm.list, readRDS)
names(cm.files) <- sub(path,"",
                       sub("\\_cell.counts.matrices.rds", "", cm.list))
```

pre-processing

```{r pre-process}
cm.pp <- mapply(EpicPreHS, cm.files, orig.ident = names(cm.files), SIMPLIFY = F)
```

Merge
```{r merge}
covid_combined.emat <- mergeCM(cm.pp, type = "emat")
covid_combined.nmat <- mergeCM(cm.pp, type = "nmat")
```

Make Seurat object
```{r create object}
covid_combined <- CreateSeuratObject(counts = covid_combined.emat, min.cells = 10, names.field = 1, names.delim = "\\.")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^MT-", col.name = "percent.mt")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^RPS", col.name = "percent.rps")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^RPL", col.name = "percent.rpl")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^RNA\\d8S5", col.name = "percent.rrna")
covid_combined <- SCTransform(covid_combined, vars.to.regress = c("percent.mt", "percent.rps", "percent.rpl", "percent.rrna", "nCount_RNA", "nFeature_RNA"), verbose = FALSE, return.only.var.genes = TRUE) #expect "iteration limit reached" warning unless suppressed per https://github.com/satijalab/seurat/issues/1426

covid_combined <- RunPCA(covid_combined, verbose = FALSE)
covid_combined <- RunUMAP(covid_combined, dims = 1:50, verbose = FALSE)
covid_combined <- FindNeighbors(covid_combined, dims = 1:50, verbose = FALSE)
covid_combined <- FindClusters(covid_combined, resolution = 1, verbose = FALSE)
DimPlot(covid_combined, label = TRUE) + NoLegend()
```

Add metadata

```{r add metadata}
covid_metadata <- read_csv("https://raw.githubusercontent.com/ajwilk/2020_Wilk_COVID/master/code/COVID-19_metadata_repo.csv")
seurat_metadata <- covid_combined@meta.data
seurat_metadata <- seurat_metadata[,!(colnames(seurat_metadata)) %in% 
                                     setdiff(colnames(covid_metadata), "orig.ident")]
metadata_combined <- merge(seurat_metadata, covid_metadata, by = "orig.ident")
rownames(metadata_combined) <- rownames(covid_combined@meta.data)

covid_combined@meta.data <- metadata_combined
```


```{r}
sars.names <- c("MN994467.1", "MN994468.1", "MT027062.1", "MT027063.1", "MT027064.1", "MT044258.1", "MT106052.1", "MT106053.1", "MT118835.1", "MT192765.1")
covid.pos <- covid_combined.emat[grep(paste(sars.names,collapse="|"), rownames(covid_combined.emat)),]
covid.pos.cells <- names(covid.pos[covid.pos !=0])
length(covid.pos.cells) #no SARS-CoV-2 reads detected in dataset
```



```{r}
covid_combined.markers <- FindAllMarkers(covid_combined, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
covid_combined.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)
```
Cluster 3 is a low quality cluster. Are there any genes other than poor quality genes that are upregulated in it?
```{r}
crap.markers <- FindMarkers(covid_combined, ident.1 = "3")
crap.markers[order(-crap.markers$avg_logFC),]
```

Will remove. Same with cluster 12
```{r}
crap2.markers <- FindMarkers(covid_combined, ident.1 = "12")
crap2.markers[order(-crap2.markers$avg_logFC),]
```
Will also remove cluster 12

SingleR annotation
```{r}
ref <- HumanPrimaryCellAtlasData()
common <- intersect(rownames(ref), rownames(covid_combined.emat))
combined.emat.singler <- covid_combined.emat[common,]
ref <- ref[common,]
combined.emat.singler.sce <- SingleCellExperiment(assays = list(counts = combined.emat.singler))
combined.emat.singler.sce <- logNormCounts(combined.emat.singler.sce)
```
```{r}
singler.pred <- SingleR(test = combined.emat.singler.sce@assays@data@listData$logcounts, ref = ref, labels = ref$label.main)
table(singler.pred$labels)
```

```{r}
plotScoreHeatmap(singler.pred, clusters = covid_combined@meta.data$orig.ident)
plotScoreHeatmap(singler.pred, clusters = covid_combined@meta.data$seurat_clusters)
```

```{r}
singler.results <- merge(data.frame(cell = rownames(singler.pred), singler = singler.pred$labels), 
                         data.frame(cell = rownames(covid_combined@meta.data), 
                                    cluster = covid_combined@meta.data$seurat_clusters), 
                         by = "cell", 
                         all.y = FALSE)
singler.results$cell <- NULL
singler.results$count <- 1
singler.results <- aggregate(count ~ ., singler.results, FUN = sum)
singler.final <- singler.results %>% group_by(cluster) %>% top_n(n = 1, wt = count)
singler.final
covid_combined$singler <- singler.pred$labels
DimPlot(covid_combined, group.by = "singler", label = T) + NoLegend()
```


Removing crap clusters, re-embedding and re-clustering
```{r}
covid_combined.nc <- subset(covid_combined, idents = setdiff(unique(covid_combined$seurat_clusters), c("3", "12")))
covid_combined.nc <- RunPCA(covid_combined.nc, verbose = FALSE)
covid_combined.nc <- RunUMAP(covid_combined.nc, dims = 1:50, verbose = FALSE)
covid_combined.nc <- FindNeighbors(covid_combined.nc, dims = 1:50, verbose = FALSE)
covid_combined.nc <- FindClusters(covid_combined.nc, resolution = 1, verbose = FALSE)
DimPlot(covid_combined.nc, label = TRUE) + NoLegend()
```


```{r}
singler.results <- merge(data.frame(cell = rownames(singler.pred), singler = singler.pred$labels), 
                         data.frame(cell = rownames(covid_combined.nc@meta.data), 
                                    cluster = covid_combined.nc@meta.data$seurat_clusters), 
                         by = "cell", 
                         all.y = FALSE)
singler.results$cell <- NULL
singler.results$count <- 1
singler.results <- aggregate(count ~ ., singler.results, FUN = sum)
singler.final <- singler.results %>% group_by(cluster) %>% top_n(n = 1, wt = count)
singler.final
```

Coarse annotations: 
0: NK cell
1: T cell
2: T cell
3: Monocyte
4: T cell
5: B cell
6: Monocyte
7: Monocyte
8: Monocyte
9: B cell
10: Monocyte
11: NK cell
12: RBC
13: T cell
14: RBC
15: T cell
16: B cell
17: Platelet
18: B cell
19: T cell
20: Monocyte
21: B cell
22: T cell
23: B cell
24: CMP
25: Neutrophil
26: Monocyte
27: Myelocyte
28: Monocyte
29: B cell


```{r}
covid_combined.nc.markers <- FindAllMarkers(covid_combined.nc, only.pos = T)
covid_combined.nc.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)
```

```{r}
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, 
                               idents = unlist(lapply(singler.final[singler.final$singler=="T_cells","cluster"],
                                                      as.character))), 
        features = c("CD4", "PRF1", "GZMB", "GZMA", "CD8A", "RORC", "EOMES", "TBX21"))
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, 
                               idents = unlist(lapply(singler.final[singler.final$singler=="T_cells","cluster"],
                                                      as.character))), 
        features = c("PTK7", "PECAM1", "ITGAE", "CD44", "CR2", "IL2RA", "FAS", "IL2RB"))
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, 
                               idents = unlist(lapply(singler.final[singler.final$singler=="T_cells","cluster"],
                                                      as.character))), 
        features = c("SELL", "CCR7", "CD44", "CD28", "CD27", "IL7R", "CXCR3", "ITGAL", "CD58"))
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, 
                               idents = unlist(lapply(singler.final[singler.final$singler=="T_cells","cluster"],
                                                      as.character))), 
        features = c("TRDC", "TRGC1", "TRGC2", "KLRB1", "KLRK1"))
```
```{r}
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, 
                               idents = unlist(lapply(singler.final[singler.final$singler=="Monocyte","cluster"],
                                                      as.character))), 
        features = c("CD14", "LYZ", "FCGR3A", "IL1B", "IDO1", "FCER1A", "FLT3", "IL3RA", "NRP1"))
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, 
                               idents = unlist(lapply(singler.final[singler.final$singler=="B_cell","cluster"],
                                                      as.character))), 
        features = c("MME", "CD22", "FCER2", "CD38", "CD44", "CD27", "SDC1"))
```

```{r fig.width=6, fig.height=3}
DotPlot(Seurat:::subset.Seurat(covid_combined.nc, idents = c(covid_myeloid.idents)), features = unique(c("CD14", "LYZ", "FCGR3B", "FCGR3A", "CLC", "ELANE", "LTF", "MPO", "CTSG", "IDO1", "FCER1A", "FLT3", "IL3RA", "NRP1", "MME", "CD22", "FCER2", "CD44", "CD27", "SDC1", "CD4", "CD8A", "ITGAL", "SELL", "GZMB", "CD3E", "CD3G", "CD3D")), group.by = "cell.type.fine") + ggpubr::rotate_x_text()
ggsave("p.pdf", path = "~/Downloads/", height = 5, width = 10)
```


Fine cluster annotations: 
0: NK cell
1: Memory CD8 T cell
2: Memory CD4 T cell
3: CD14 Monocyte
4: Naive CD4 T cell
5: B cell
6: CD14 Monocyte
7: CD14 Monocyte
8: CD14 Monocyte
9: Class-switched B cell
10: CD16 Monocyte
11: NK cell
12: RBC
13: Effector CD8 T cell
14: RBC
15: Memory CD8 T cell
16: IgG Plasmablast
17: Platelet
18: IgG Plasmablast
19: CD4 T cell
20: DC
21: IgA Plasmablast
22: gd T cell
23: IgA Plasmablast
24: CMP/Eosinophil
25: Neutrophil
26: pDC
27: Myelocyte/Neutrophil
28: CD16 Monocyte
29: IgA Plasmablast

Myeloid compartment clusters: 3, 6, 7, 8, 10, 20, 24, 25, 26, 27
B cell clusters: 5, 9, 16, 18, 21, 24, 27, 29
T cell clusters: 1, 2, 4, 13, 15, 19, 22
NK cell clusters: 0, 11

```{r}
covid_myeloid.idents <- c("3", "6", "7", "8", "10", "20", "24", "25", "26", "27", "28")
covid_B.idents <- c("5", "9", "16", "18", "21", "24", "27", "29")
covid_T.idents <- c("1", "2", "4", "13", "15", "19", "22")
covid_NK.idents <- c("0", "11")

covid_CD4T.idents <- c("2", "4", "19")
covid_CD8T.idents <- c("1", "13", "15")
covid_gdT.idents <- c("22")
covid_CD14mono.idents <- c("3", "6", "7", "8")
covid_CD16mono.idents <- c("10", "28")
covid_DC.idents <- c("20", "26")

covid_fine.idents <- c("NK", "CD8m T", "CD4m T", "CD14 Monocyte", "CD4n T", "B", "CD14 Monocyte", "CD14 Monocyte", "CD14 Monocyte", "IgM PB", "CD16 Monocyte", "NK", "RBC", "CD8eff T", "RBC", "CD8m T", "IgG PB", "Platelet", "IgG PB", "CD4 T", "DC", "IgA PB", "gd T", "IgA PB", "SC & Eosinophil", "Neutrophil", "pDC", "Developing Neutrophil", "CD16 Monocyte", "IgA PB")

covid_coarse.idents <- c("NK", "CD8 T", "CD4 T", "CD14 Monocyte", "CD4 T", "B", "CD14 Monocyte", "CD14 Monocyte", "CD14 Monocyte", "B", "CD16 Monocyte", "NK", "RBC", "CD8 T", "RBC", "CD8 T", "PB", "Platelet", "PB", "CD4 T", "DC", "PB", "gd T", "PB", "Granulocyte", "Granulocyte", "pDC", "Granulocyte", "CD16 Monocyte", "PB")
```


This Seurat object can now be used for post-processing analysis using the "covid_analysis_markdown.Rmd" notebook.
